\documentclass{article}

\title{Proposal and Contract for 4AL3 ML Project}
\author{Rebecca Di Filippo, Jake Read, Claire Nielsen}
\date{\today}

\begin{document}

\maketitle

\newpage
\section{Team Members}

\textbf{Rebecca Di Filippo}\\
difilir@mcmaster.ca\\

\textbf{Jake Read}\\
readj9@mcmaster.ca\\

\textbf{Claire Nielsen}\\
nielsc2@mcmaster.ca\\

\section{Task Title and Overview}

\section{Task Definition}
(Type of data, classification/regression/generation, number of classes, single-label or multi-label\\

\section{Problem Description}
Describe the problem, it’s impact on the real world, and why it is challenging \\

\section{Data Source(s) and Plan for Data Collection.}
This may include how you are going to scrape the data and follow terms-of-service, API access and handling rate limiting, using open-source data, or any other relevant details. If your data does not have labels, how do you plan to get them? If assigning labels by hand, how long does it take per instance? Include links to the data if relevant. If you download your dataset from Kaggle, you must include the Kaggle link and the original data source link from where the dataset was downloaded and posted on
Kaggle. Include any meta-data available for your corpus. If you are not collecting it yourself, include details about how the data was collected, annotated, preprocessed, etc. Do not work on the datasets for which no source information is available. Please indicate whether you will use a small subset of the data or features for your project or the entire dataset, and why. \\

\section{Expected Size of the Dataset}
(number of data points) and 3 example data points with labels. Your dataset should have at least 1k data points. Some projects will have more or less data. \\

\section{Proposed Solution}
How do you plan to go about solving this problem? It is okay to not know
how the machine learning models work at this point in the class, but you should be starting to get some idea based on the lectures and assigned readings. What kind of features and target labels do you have? What kind of models might you try? You may not propose simple linear regression models for this project. Are there any existing solutions to your problem?
You must indicate 2-5 sources (research papers, books, machine learning challenges online) from where your solution was inspired. Put some thought into how you would approach this. How will you know if the model is good? How will you evaluate it? Also, share the libraries you intend to use for the project. \\

\newpage
\section{Team Charter}
A team contract signed by all team members. This is a description of the expectations and/or roles that team members agree upon for the semester. Part of your final project grade will be based on whether your teammates agree that you followed this contract. See next section for details.

\end{document}
